What is an Ancestral Recombination Graph?

### 2022-02-15 paper.tex
Very simple GARG definitions

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
There are many ways in which the EARG structure can be formalised (see
e.g.~\citep{gusfield2014recombinatorics}).
% It would be slightly easier for us if we defined nodes as in N_0, since
% we use 0 as a node ID, but leaving for now.
We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines a directed edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
% Because of our definitions we can't have anything to the left of 0,
% so you can treat breakpoint at 0 as no recombination.
$\sigma(u) = 0$ otherwise. For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

Given a set of sample nodes $S$ we can recover the local tree at
position $x$ by the following procedure. Begin by initialising
a stack

@@

Formally, we can define a GARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r])$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).

@@

We can define a GARG as a set of directed edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$.

We can define a sample-resolved GARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r])$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.
% Should probably define how we build the local trees somewhere.

>>>>

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
There are many ways in which the EARG structure can be formalised (see
e.g.~\citep{gusfield2014recombinatorics}).
% It would be slightly easier for us if we defined nodes as in N_0, since
% we use 0 as a node ID, but leaving for now.
We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines a directed edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
% This is arbitrary, but handy when you start defining the function
% for recovering trees; see below
$\sigma(u) = \infty$ otherwise. For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

% Not sure if we actually include this, but it's good to work it through
% to make sure the notation works.
Given a set of sample nodes $S$ we can recover the local tree at
position $x$ by the following procedure. Let $N$ be a set of
nodes we are currently constructing the tree at, and let
$p: \mathbb{N} \rightarrow \mathbb{N}$ be the mapping from child
node to parent. Firstly, initialise $N \leftarrow S$.
Then, while $|N| > 0$, we add branches to the
tree as follows. Let $u$ be an arbitrary element of $N$ (which
we remove). Then, if $\sigma(u) < x$ we let $j$ be the
smallest value such that $e_{j,1} = u$, and otherwise
let $j$ be the largest value such that $e_{j,1} = u$.
Then, set $v \leftarrow e_{j, 2}$,
$p_u \leftarrow v$
and $N \leftarrow N \cup \{v\}$.

@@

Formally, we can define a GARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).

@@

We can define a sample-resolved GARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.
% Should probably define how we build the local trees somewhere.
